Studying public transit API query logs\\ to get an indication of travel flows

### 2016-01-02 ldfaxis2.tex
abstract and wrote publishing transport data

    \draw[axis] (-100, 0) -- (-5, 0);
    \node[axislabel,align=left,anchor=north west] at (-100, -2.3)
    {
%      generic requests\\
      high client effort\\
      high server availability\\
      know how many times datadump was downloaded
    };
    \node[axislabel,align=right,anchor=north east] at (-5, -2.3)
    {
%      specific requests\\
      high server effort\\
      low server availability\\
      detailled query logs
    };
    \node[text,ticklabel] at (-53, 2.4) {Linked\\Connections};
    \tick{-90}{GTFS data\\dump}
    \tick{-15}{route planning\\service}
    \vaguetick{-23}
    \vaguetick{-32}
    \vaguetick{-41}
    \vaguetick{-50}
    \vaguetick{-60}
    \vaguetick{-70}
    \vaguetick{-80}

   %\node[align=center,font=\fontsize{8}{9}\selectfont\bfseries] at (0, -8.3)
   %   {various types of\\Linked Data Fragments};
  \end{tikzpicture}
  \caption{This axis extends \cref{fig:LDFAxis1} with Linked Connections: the different ticks on the axis for client effort versus server expressivity.}
  \label{fig:LDFAxis2}
\end{figure}

>>>>

    \draw[axis] (-100, 0) -- (-5, 0);
    \node[axislabel,align=left,anchor=north west] at (-100, -2.3)
    {
%      generic requests\\
      high client effort\\
      high server availability\\
      know how many times datadump was downloaded
    };
    \node[axislabel,align=right,anchor=north east] at (-5, -2.3)
    {
%      specific requests\\
      high server effort\\
      low server availability\\
      detailed query logs
    };
    \node[text,ticklabel] at (-53, 2.4) {Linked\\Connections};
    \tick{-90}{GTFS data\\dump}
    \tick{-15}{route planning\\service}
    \vaguetick{-23}
    \vaguetick{-32}
    \vaguetick{-41}
    \vaguetick{-50}
    \vaguetick{-60}
    \vaguetick{-70}
    \vaguetick{-80}

   %\node[align=center,font=\fontsize{8}{9}\selectfont\bfseries] at (0, -8.3)
   %   {various types of\\Linked Data Fragments};
  \end{tikzpicture}
  \caption{This axis, first introduced by Linked Data Fragments~\cite{ldf}, illustrates the possibilities within publishing transport data. Different ticks on the axis illustrate client effort versus server expressivity: on the far left, data dumps offer high server availability, yet the effort needed by data reusers is high, and query logs do not reach the server. Moving to the right, we identify Linked Connections~\cite{lc} as an in-between solution. On the far right, route planning services require high server effort, leading to detailed query logs.}
  \label{fig:LDFAxis2}
\end{figure}

### 2016-01-02 paper.tex
abstract and wrote publishing transport data

%Public transit schedules are made available on the Web in various ways, such as in a GTFS file, a route planning API, or through Linked Connections.
%This webservice provides an answer to a route planning question with parameters such as ``departuretime'', ``from'' and ``to''.
% Context & % Need
In the field of urban planning, researchers need an indication of how people move between cities. 
Yet, getting statistics of travel flows within public transit systems has proven to be troublesome.
% Task
We analysed the query logs of the iRail API, a high expressive route planning API for the Belgian railways, to get an indication of travel flows between cities in Belgium.
We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which is between $0.5\%$ and $1.7\%$ of the amount of monthly passengers.
% Object & Findings
Using data visualizations, we illustrated the commuting patterns in Belgium and show that Brussels, the capital, acts as a central hub. The Flemish region appears to be polycentric, while in the Walloon region, everything converges on Brussels.
The findings correspond to the real travel demand, say experts of the passenger federation Trein Tram Bus.
% Conclusion
We concluded that query logs of route planners are of high importance in getting an indication of the travel flows.
% Perspectives
\todo{another sentence about publishing transit data for maximized reuse and how to gather analytics}

@@

We cleaned the logs by removing the requests done by harvesters or search engine bots.
Thanks to the user agents, this was a fairly straight-forward operation.
The remainder of the file delivers an average of 3000 route planning queries a day, which we can use an an indication of travel intensities.
Taking into account that 758,370 people take the train each day on a weekday (according to official number of the Belgian Railway company in 2013\footnote{\url{http://www.treintrambus.be/images/OpstappendeReizigers_2013_2.pdf}}) and assuming that each query is a trip done by a certain passenger, this sample equals $\sim0.4\%$ of the railway users in Belgium.

@@

It is in the interest of public transit agencies that the use of their time schedule data is maximized.
We can imagine that when more people are informed about the agency's offer, there will be a higher amount of people that will use the service.
There are currently two ways to publish transit data, as illustrated in \cref{fig:LDFAxis2}: publishing data using the General Transit Feed Specification (GTFS) and offering a route planning service.

\emph{GTFS}, on the one hand, is a data dump format: it is a compressed ZIP-file, containing a couple of CSV files, describing the rules for when a public transit vehicle will pass by on a certain location.
GTFS is supported among all current open source route planning software systems and it is used in products/apps such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
%Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
The goal of GTFS is to create an exchange format which specific software packages can use to transform it into their format of choice.

On the other hand, public transit agencies also publish their data by providing \emph{route planners} such as the French railway company\footnote{\url{http://tech.eu/features/7119/sncf-api-real-time-data/}}, the Dutch railway company\footnote{\url{http://www.ns.nl/reisinformatie/ns-api}} or \ldots\footnote{\url{http://otherapiofadifferentagency}}
These route planner offer a query service on top of the data and expose these over the Web.
In this paper we have studied the query logs of such an \emph{expressive} route planning API, which per request offers an interesting log entry.

@@

Linked Data Fragments~\cite{ldf} is a way to think about data publishing as a set of trade-offs between client and servers.
These trade-offs can be visualized on the axis in \cref{fig:LDFAxis2}: ticks in between the extremes indicate different client/server options.
For route planning systems, we call these other ticks Linked~Connections~\cite{lc}: 
by publishing fragments of the data needed by an earliest arrival time algorithm, the algorithm can be executed by a user agent while the downloading the data.
An example of such an implementation is illustrated in \cref{fig:lc}.
By exploiting the caching mechanisms behind HTTP, it lowers to load on the server, making it easier to keep the data high available.
Furthermore, it enables clients to federate queries over different servers representing different regions, transit modes or different user requirements such as low criminality rates or wheelchair accessibility.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.48\textwidth]{LC}
    \caption{An example implementation of a route planning algorithm: fragments of data needed by a route planning algorithm are published in pages. The route planning algorithm will need multiple requests instead of just one. The fragements of the pages that can be retrieved are however highly cacheable.}
    \label{fig:lc}
\end{figure}

The expressiveness of the server also affects the way logs can be gathered.
If we want to log every HTTP GET request, we expect every HTTP GET to reach the server and we would have to deliberately disable caches to be able to log every request.
The chance that two requests are exactly the same for a high expressive server is low, as there are more URLs which can be requested. 
We have shown in previous sections that these logs are interesting: each request contains an entire query which can be used as an indication of travel ``demand''.

A low expressive server that still publishes the data in a queryable way as illustrated in \cref{fig:lc}, has less interesting results. 
E.g., when it only allows to set a departure station and a departure time, then the server cannot know the arrival station, which makes it harder to get an indication of the travel flow.
Furthermore, on the Linked Data Fragments axis, less expressive servers --which have less URLs to be requested-- can exploit the caching mechanism~\cite{ldf}.
The drawback of having higher cacheability, is that intermediate proxies may have cached the HTTP response somewhere in between and there will not be a log entry.

When the expressiveness of the servers is lower in order to maximize the reuse of the data, such as in the case of Linked~Connections or a GTFS~datadump, we suggest to, instead of logging each GET-request, to have an analytics server that gathers all the logs.

\todo{\ldots}

@@

%\subsection{Intermodal route planning}

As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
This amounts today to an average of 86k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\url{https://play.google.com/store/apps/details?id=tof.cv.mpp}} and \emph{Railer}\footnote{\url{http://railer.be/}}.
On average, 14,357 of these are valid route planning queries (other queries include requesting departures in a station, status of a vehicle, or a list of all stations).
If each query would amount to indeed one passenger, this would today be equal to almost 2\% of all passengers.

% Open Anonymized Route Planning Query Logs
Semantics behind the query logs

>>>>

%Public transit schedules are made available on the Web in various ways, such as in a GTFS file, a route planning API, or through Linked Connections.
%This webservice provides an answer to a route planning question with parameters such as ``departuretime'', ``from'' and ``to''.
% Context & % Need
In the field of urban planning, researchers need an indication of how people move between cities. 
Yet, getting statistics of travel flows within public transit systems has proven to be troublesome.
% Task
We analyzed the query logs of the iRail API, a high expressive route planning API for the Belgian railways, to get an indication of travel flows between cities in Belgium.
We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which is between $0.5\%$ and $1.7\%$ of the amount of monthly passengers.
% Object & Findings
Using data visualizations, we illustrated the commuting patterns in Belgium and show that Brussels, the capital, acts as a central hub. The Flemish region appears to be polycentric, while in the Walloon region, everything converges on Brussels.
The findings correspond to the real travel demand, say experts of the passenger federation Trein Tram Bus.
% Conclusion
We conclude that query logs of route planners are of high importance in getting an indication of the travel flows.
% Perspectives
High expressive transport data publishing methods such as route planning API exist, as well as low expressive data dumps or data fragments.
In order to be able to gather meaningful logs in all cases, we suggest using a separate POST request containing the entire query.

@@

We cleaned the logs by removing the requests done by harvesters or search engine bots.
Thanks to the user agents, this was a fairly straight-forward operation.
\todo{The remainder of the file delivers an average of 3000 route planning queries a day, which we can use an an indication of travel intensities.
Taking into account that 758,370 people take the train each day on a weekday (according to official number of the Belgian Railway company in 2013\footnote{\url{http://www.treintrambus.be/images/OpstappendeReizigers_2013_2.pdf}}) and assuming that each query is a trip done by a certain passenger, this sample equals $\sim0.4\%$ of the railway users in Belgium.}\footnote{Calculated by using indicators of the Flemish government and extrapolating the numbers for 2015 \url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}

@@

%It is in the interest of public transit agencies that the use of their time schedule data is maximized.
%We can imagine that when more people are informed about the agency's offer, there will be a higher amount of people that will use the service.
%There are currently two ways to publish transit data, as illustrated in \cref{fig:LDFAxis2}: publishing data using the General Transit Feed Specification (GTFS) and offering a route planning service.

%\emph{GTFS}, on the one hand, is a data dump format: it is a compressed ZIP-file, containing a couple of CSV files, describing the rules for when a public transit vehicle will pass by on a certain location.
%GTFS is supported among all current open source route planning software systems and it is used in products/apps such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
%Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
%The goal of GTFS is to create an exchange format which specific software packages can use to transform it into their format of choice.

%On the other hand, public transit agencies also publish their data by providing \emph{route planners} such as the French railway company\footnote{\url{http://tech.eu/features/7119/sncf-api-real-time-data/}}, the Dutch railway company\footnote{\url{http://www.ns.nl/reisinformatie/ns-api}} or \ldots\footnote{\url{http://otherapiofadifferentagency}}
%These route planner offer a query service on top of the data and expose these over the Web.
%In this paper we have studied the query logs of such an \emph{expressive} route planning API, which per request offers an interesting log entry.

@@

%Linked Data Fragments~\cite{ldf} is a way to think about data publishing as a set of trade-offs between client and servers.
%These trade-offs can be visualized on the axis in \cref{fig:LDFAxis2}: ticks in between the extremes indicate different client/server options.
%For route planning systems, we call these other ticks Linked~Connections~\cite{lc}: 
%by publishing fragments of the data needed by an earliest arrival time algorithm, the algorithm can be executed by a user agent while the downloading the data.
%An example of such an implementation is illustrated in \cref{fig:lc}.
%By exploiting the caching mechanisms behind HTTP, it lowers to load on the server, making it easier to keep the data high available.
%Furthermore, it enables clients to federate queries over different servers representing different regions, transit modes or different user requirements such as low criminality rates or wheelchair accessibility.

%\begin{figure}[h]
%    \centering
%    \includegraphics[width=0.48\textwidth]{LC}
%    \caption{An example implementation of a route planning algorithm: fragments of data needed by a route planning algorithm are published in pages. The route planning algorithm will need multiple requests instead of just one. The fragements of the pages that can be retrieved are however highly cacheable.}
%    \label{fig:lc}
%\end{figure}
%

%%% EXPLAIN:
%%%
The expressiveness of a server affects the way logs can be gathered.
We have shown in previous sections that the query logs of a \emph{high expressive server}, such as the iRail API, are interesting: each request contains an entire query which can be interpreted as a travel intention.
The possibility that two requests are exactly the same is low, as there are many URLs which can be requested.
Nevertheless, we cannot fully guarantee that each HTTP GET request to the server will trigger a log entry, due to caching mechanisms on the Web~\cite{fielding}, which might lead to a false representation.

%intermodality, features and availability
Hosting a route planning API as the only way to publish transport data comes with three identified limitations, as the server will need to handle the requests from different use cases with different needs:
\begin{enumerate}
  \item When an application developer would like a \emph{new feature}, such as taking wheelchair accessibility information into account, the feature would have to be implemented on the server of the data publisher
  \item Keeping the server \emph{high available} is costly, as any question can be asked by anyone for any purpose.
  \item Federated querying, which would allow for \emph{intermodal} route planning for route planning APIs, is unexistent up to date.
\end{enumerate}

% GTFS
In order to overcome these limitations, the General Transit Feed Specification (GTFS)\footnote{\url{https://developers.google.com/transit/gtfs/reference}} can be used. 
GTFS is a compressed ZIP-file containing a couple of CSV files, describing the rules for when a public transit vehicle will pass by on a certain location.
It is supported among all current open source route planning software systems and it is used in products/apps such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
%Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
%The goal of the format is to  an exchange format which specific software packages can use to transform it into their format of choice.
It succesfully enabled reuse for intermodal travel, engineers can rely on the data dumps even if the servers of the transit agency are offline and there is no limitation to the features that can be implemented.
The logs are however lost.

In \cref{fig:LDFAxis2} we illustrate these two options as two extremes, with other options that are yet to be discovered.
When a \emph{rather low expressive} server only allows to set e.g., a departure station and a departure time, then the server cannot log the arrival station, yet the client is still able to plan a route by executing the algorithm on the client-side~\cite{lc}.
Again, we are not able to rely on the query logs

\todo{When the expressiveness of the servers is lower in order to maximize the reuse of the data, such as in the case of Linked~Connections or a GTFS~datadump, we suggest to, instead of logging each GET-request, to have an analytics server that gathers all the logs.}

@@

\todo{As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
This amounts today to an average of 86k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\url{https://play.google.com/store/apps/details?id=tof.cv.mpp}} and \emph{Railer}\footnote{\url{http://railer.be/}}.
On average, 14,357 of these are valid route planning queries (other queries include requesting departures in a station, status of a vehicle, or a list of all stations).
If each query would amount to indeed one passenger, this would today be equal to almost 2\% of all passengers.}
